Beneficial arthropod habitat useage changes within-season: implications for pest control services in croplands

### 2020-10-02 manuscript/manuscript_outline.Rmd
Made plots for box in manuscript

## Data collection
We used a set of `r sum(NtrapsPer$n)` pitfall traps installed in road margins (minimum of 5 m away from the road edge, `r with(NtrapsPer, sum(n[trapLoc=='ditch']))` traps) and in-field locations (`r with(NtrapsPer, sum(n[trapLoc!='ditch']))`) across southern Alberta, Canada in 2017 (Figure \@ref(fig:siteMap)).
Traps were placed starting on `r format(as.Date(as.character(min(PteMelMod$tempTrap$startjulian)),format='%j'),format='%B %d')`, and collections ended on `r format(as.Date(as.character(max(PteMelMod$tempTrap$endjulian)),format='%j'),format='%B %d')`, with collection occurring continuously and traps being emptied every `r round(mean(PteMelMod$tempTrap$trapdays),1)` days on average (SD: `r round(sd(PteMelMod$tempTrap$trapdays),1)`). 
This resulted in `r nrow(PteMelMod$tempTrap)` unique collection events, taking place across a total of `r round(sum(PteMelMod$tempTrap$trapdays),1)` trapping days.
The sites spanned a west-to-east gradient of four natural subregions, including foothills parkland, foothills fescue, mixedgrass, and dry mixedgrass [@ABRegions2006].
In-field traps were placed in canola crops (`r with(NtrapsPer, sum(n[trapLoc=='canola']))`), wetlands (`r with(NtrapsPer, sum(n[trapLoc=='wetland']))`), grassy field margins (`r with(NtrapsPer, sum(n[trapLoc=='pivot']))`), and remnant prairie grasslands (`r with(NtrapsPer, sum(n[trapLoc=='native']))`).
Traps in canola were installed at 25, 75, and 200 m along a transect heading away from the nearest non-crop feature, while the trap at 0 m was installed in the non-crop feature itself.
We used 582 mL Solo® cups buried up to the rim and partially filled with ethylene glycol, with 2 cm wire mesh mounted over the rim to prevent vertebrates from falling into the traps.
Specimens were identified to species using appropriate taxonomic literature [@dondale1990; @edgar1990; @vogel2004; @yigit2007].

@@

```{r siteMap, echo=FALSE, fig.cap='Location of pitfall traps during 2017, and classified cover f. Inset map shows study location within Alberta. Roadside traps were installed in road margins, while in-field traps were installed directly in canola fields or in adjacent wetland or grassy field corners.', out.width='100%'}  
  include_graphics('../figures/maps/PF2017Map.png',dpi=NA)
```

We used four highly-abundant species of predatory arthropods found in the pitfall traps, as common species are often more important for ecosystem service provision than rare ones [@winfree2015].
_Pterostichus melanarius_ is an introduced predatory ground beetle (Coleoptera: Carabidae) that is a wide-ranging generalist [@carcamo1994; @larsen2003; @busch_Msc].
It is commonly found in untilled grasslands [@purtauf2005] and may use grassy field margins as larval habitat [@desender1988].
_Pardosa_ are wolf spiders (Araneae: Lycosidae) that use an active wandering predation strategy [@young1990], and are common across the Canadian prairies.
Both _Pardosa distincta_ and _Pardosa moesta_ are found in a wide variety of habitats, but are common in ungrazed pastures [@dondale1990; @carcamo2014].
_P. distincta_ is commonly found in disturbed environments [@collins1996; @wade2010], while _P. moesta_ appear to prefer less frequent disturbances [@graham2003; @kowal2011; but see @dondale1990].
Other _Pardosa_ are known to use road margins and wooded areas as habitat [@buddle2000; @drapela2011].
Finally, _Phallangium opilio_ is an introduced harvestman (Opiliones: Phalangiidae) that is found in drier disturbed areas [@bragg2009].
Juveniles hatch from eggs during the spring, becoming adults during the summer, and eggs are laid during the fall [adults do not overwinter; @bragg2009].
They are commonly found in human-altered landscapes [@muster2014; @vandepoel2015Msc] and are nocturnal generalist hunters and scavengers [@halaj2000; @allard2005; @allard2005b], mainly eating small soft-bodied invertebrates [@acosta2007].

@@

Scalar-on-function regression of activity-density was fit using generalized additive models [_mgcv_ version `r packageVersion('mgcv')`; @wood2017]. 
Count data of arthropods were modeled using a negative binomial distribution with a log-link function and a single dispersion parameter ($\theta$).
To account for different lengths of trap exposure, log-days since trap placement were used as a fixed effect with their slope held at 1 ("offset" variable).
We included day of year as a 1-dimensional smooth, and easting and northing (km east and north of the trap extent centre point) as a 2-dimensional smooth, in order to account for underlying spatial and temporal autocorrelation in the data.
The deviance residuals from each model were visually inspected for normality and equal variance [@hilbe2011; @wood2017].

@@

@@

Grassland, cereal, and canola were the three most-abundant landcover classes surrounding our traps, accounting for `r coverList %>% group_by(cover) %>% summarize(p=sum(prop)) %>% ungroup() %>% arrange(desc(p)) %>% mutate(prop=round(100*p/sum(p),2)) %>% slice(1:3) %>% pull(prop) %>% sum `% of land cover (Figure \@ref(fig:landscapeComp)).
Several landscape ``fingerprints" were evident in the landscape annuli, with cereal cover increasing with distance away, along with a corresponding decrease in canola cover.
Grassland cover was largely constant with distance, although a cluster of sites had uniformly high or low cover of grassland with distance.
The pitfall traps caught a total of `r with(Ncatches,total[spp=='PteMel'])` _Pterostichus melanarius_, `r with(Ncatches,total[spp=='ParDis'])` _Pardosa distincta_, `r with(Ncatches,total[spp=='ParMoe'])` _Pardosa moesta_, and `r with(Ncatches,total[spp=='Opilio'])` _Phalangium opilio_ (mean trapping rates per day: `r with(Ncatches,paste(round(perDayMean,2),collapse=', '))`, SDs: `r with(Ncatches,paste(round(perDaySD,2),collapse=', '))`).

@@

<!-- Percent signs are escaped in LaTeX, but this causes havoc if used in fig.cap of chunk options (OK if used in internal captions, I think). Yihui uses "text references" instead. -->
(ref:PteMelCap) Landscape influence on _Pterostichus melanarius_ activity-density. Lines and dots represent means, and bars and shaded regions represent 95 % confidence intervals (1.96 x SE). Coloured regions represent early-, mid-, and late-season effects (red, green, blue).

@@

(ref:ParDisCap) Landscape influence on _Pardosa distincta_ activity-density. Lines and dots represent means, and bars and shaded regions represent 95 % confidence intervals (1.96 x SE). Coloured regions represent early-, mid-, and late-season effects (red, green, blue).

@@

(ref:ParMoeCap) Landscape influence on _Pardosa moesta_ activity-density. Lines and dots represent means, and bars and shaded regions represent 95 % confidence intervals (1.96 x SE). Coloured regions represent early-, mid-, and late-season effects (red, green, blue).

@@

_Phalangium opilio_ activity-density was strongly influenced by trap location, with pivot corners and wetlands having the highest activity-density (Figure \@ref(fig:OpilioLandscape)a).
Landscape composition explained `r filter(r2Nakagawa2,grepl('Opilio',model),!grepl('s\\(',terms),terms!='trapLoc',terms!='sigmaD') %>% summarize(s=sum(r2comp)) %>% pull(s) %>% paste(.,collapse=', ')`% of variance in _P. opilio_ activity-density.
Nearby grassland (p=`r with(opilSmTerms,p[Smoother=='ti(Distance,Day):GrassWetland'])`) and woodland (p=`r with(opilSmTerms,p[Smoother=='s(Distance):TreeShrub'])`) both had negative effects on _P. opilio_ activity-density, but only early in the season (Figures \@ref(fig:OpilioLandscape)b,c).
Compare this to _P. melanarius_, which is also influenced by grassland and woodland, but where the effect was negative _late_ in the season.
The temporal and spatial components, as in the other two species, were both very strong (both p`r with(opilSmTerms,p[grepl('E,N',Smoother)])`).
The landscape composition and trap location explained `r r2Nakagawa$OpilioMod[1]*100`%  of variance in activity-density, while the spatial and temporal smoothers accounted for `r with(r2Nakagawa,(OpilioMod[2]-OpilioMod[1])*100)`%.

(ref:OpilioCap) Landscape influence on _Phalangium opilio_ activity-density. Lines and dots represent means, and bars and shaded regions represent 95 % confidence intervals (1.96 x SE). Coloured regions represent early-, mid-, and late-season effects (red, green, blue).

@@

r2Nakagawa2 %>% select(model,terms,r2comp) %>% mutate(terms=as.character(terms),r2comp=r2comp*1) %>%
  pivot_wider(names_from=model,values_from=r2comp) %>%
  mutate(terms=case_when(terms=='trapLoc' ~ 'Trap Location', terms=='s(day)' ~ 'Day of Year',
                         terms=='s(E,N)' ~ 'Spatial Location', terms=='GrassWetland' ~ 'Grass/Wetland',
                         terms=='TreeShrub' ~ 'Trees/Shrubs', terms=='sigmaD' ~ 'Residual',
                         terms=='Urban' ~ 'Road margins', TRUE ~ terms)) %>%
  kable(.,format='latex',caption='R$^2_{GLMM}$ for model components, representing the proportion of total variance attributable to each term.',
        col.names=linebreak(x=c('Term','\\emph{Pterostichus}\n\\emph{melanarius}','\\emph{Pardosa}\n\\emph{distincta}',
                              '\\emph{Pardosa}\n\\emph{moesta}','\\emph{Phalangium}\n\\emph{opilio}'),align='l'),
        digits=c(1,3,3,3,3),align='rllll',escape=FALSE)
# test_data <- data.frame(Mean = runif(5, 3.71, 7.72),
#                         N = sample(57:59, 5, replace = TRUE),
#                         sd = c(1, rep(0, 4)),
#                         d = rep(1, 5),
#                         naod = sample(1:4, 5, replace = TRUE),
#                         a = sample(5:12, 5, replace = TRUE),
#                         sa = sample(37:44, 5, replace = TRUE))
# 
# kable(test_data,"latex" ,booktabs=T, align="c",
#       col.names=linebreak(c('Mean','\\textit{N}' ,'Strongly\n Disagree','Disagree','Neither Agree\n or Disagree','Agree','Strongly\n Agree')),
#       row.names = T,escape=F) %>%
#   row_spec(0,align = "c")

@@

_Phalangium opilio_ activity-density responded to grass and wetland cover, as well as trees and shrubs, but the trap location and landscape explained only `r r2Nakagawa$OpilioMod[1]*100`% of the variance.
Other studies have found that _P. opilio_ is mostly found in human-altered landscapes with a large proportion of farming [@muster2014; @vandepoel2015Msc].
Since most of our study area was in or near farmed land, the lack of variation is understandable, but there were a few notable landscape-level patterns that emerged.
_P. opilio_ activity-density was higher in grassy pivot corners and wetlands, but grasslands and wetlands acted as a local sink of _P. opilio_ until late in the season.
These indicate that while grass and wetlands could act as a reservoir, spillover into other land cover types may be limited, as _P. opilio_ avoids heavily grazed areas, possibly due to low humidity [@sajna2011].
_P. opilio_ is also found in large numbers at the edges of forests, and may migrate from forests into farmlands [@vandepoel2015Msc].
We found the opposite pattern: forests act as a weak sink for _P. opilio_, at least early in the year.
Unlike the other arthropods, most of the variation in activity-density for _P. opilio_ was explained by day of year (R$^2$ = `r round(filter(r2Nakagawa2,model=='OpilioMod',grepl('day',terms))$r2comp,2)`), indicating strong seasonal emergence cues.
<!-- Feels like this paragraph ends too soon. Why was there so little variation due to landscape? Link to biology of the critter. -->

>>>>

## Data collection
We used a set of `r sum(NtrapsPer$n)` pitfall traps installed in road margins (minimum of 5 m away from the road edge, `r with(NtrapsPer, sum(n[trapLoc=='ditch']))` traps) and in-field locations (`r with(NtrapsPer, sum(n[trapLoc!='ditch']))`) across southern Alberta, Canada in 2017 (Figure \@ref(fig:siteMap)).
Traps were placed starting on `r format(as.Date(as.character(min(PteMelMod$tempTrap$startjulian)),format='%j'),format='%B %d')`, and collections ended on `r format(as.Date(as.character(max(PteMelMod$tempTrap$endjulian)),format='%j'),format='%B %d')`, with collection occurring continuously and traps being emptied every `r round(mean(PteMelMod$tempTrap$trapdays),1)` days on average (SD: `r round(sd(PteMelMod$tempTrap$trapdays),1)`). 
This resulted in `r nrow(PteMelMod$tempTrap)` unique collection events, taking place across a total of `r round(sum(PteMelMod$tempTrap$trapdays),1)` trapping days.
The sites spanned a west-to-east gradient of four natural subregions, including foothills parkland, foothills fescue, mixedgrass, and dry mixedgrass [@ABRegions2006].
In-field traps were placed in canola crops (`r with(NtrapsPer, sum(n[trapLoc=='canola']))`), wetlands (`r with(NtrapsPer, sum(n[trapLoc=='wetland']))`), grassy field edges (`r with(NtrapsPer, sum(n[trapLoc=='pivot']))`), and remnant prairie grasslands (`r with(NtrapsPer, sum(n[trapLoc=='native']))`).
Traps in canola were installed at 25, 75, and 200 m along a transect heading away from the nearest non-crop feature (wetland, grassy field edges, or remnant prairie), while the trap at 0 m was installed in the non-crop feature itself.
We used 582 mL Solo® cups buried up to the rim and partially filled with ethylene glycol, with 2 cm wire mesh mounted over the rim to prevent vertebrates from falling into the traps.
Specimens were identified to species using appropriate taxonomic literature [@dondale1990; @edgar1990; @vogel2004; @yigit2007].

@@

```{r siteMap, echo=FALSE, fig.cap='Location of pitfall traps during 2017, and classified cover f. Inset map shows study location within Alberta. Roadside traps were installed in road margins, while in-field traps were installed directly in canola fields or in adjacent wetland or grassy field edges.', out.width='100%'}  
  include_graphics('../figures/maps/PF2017Map.png',dpi=NA)
```

We used four highly-abundant species of predatory arthropods found in the pitfall traps, as common species are often more important for ecosystem service provision than rare ones [@winfree2015].
_Pterostichus melanarius_ is an introduced predatory ground beetle (Coleoptera: Carabidae) that is a wide-ranging generalist [@carcamo1994; @larsen2003; @busch_Msc].
It is commonly found in untilled grasslands [@purtauf2005] and may use grassy field edges as larval habitat [@desender1988].
_Pardosa_ are wolf spiders (Araneae: Lycosidae) that use an active wandering predation strategy [@young1990], and are common across the Canadian prairies.
Both _Pardosa distincta_ and _Pardosa moesta_ are found in a wide variety of habitats, but are common in ungrazed pastures [@dondale1990; @carcamo2014].
_P. distincta_ is commonly found in disturbed environments [@collins1996; @wade2010], while _P. moesta_ appear to prefer less frequent disturbances [@graham2003; @kowal2011; but see @dondale1990].
Other _Pardosa_ are known to use road margins and wooded areas as habitat [@buddle2000; @drapela2011].
Finally, _Phallangium opilio_ is an introduced harvestman (Opiliones: Phalangiidae) that is found in drier disturbed areas [@bragg2009].
Juveniles hatch from eggs during the spring, becoming adults during the summer, and eggs are laid during the fall [adults do not overwinter; @bragg2009].
They are commonly found in human-altered landscapes [@muster2014; @vandepoel2015Msc] and are nocturnal generalist hunters and scavengers [@halaj2000; @allard2005; @allard2005b], mainly eating small soft-bodied invertebrates [@acosta2007].

@@

Scalar-on-function regression of activity-density was fit using generalized additive models [_mgcv_ version `r packageVersion('mgcv')`; @wood2017]. 
Count data of arthropods were modeled using a negative binomial distribution with a log-link function and a single dispersion parameter ($\theta$).
To account for different lengths of trap exposure, log-days since trap placement were used as a fixed effect with their slope held at 1 ("offset" variable).
Trapping location was included as a fixed effect with 5 levels (canola, field edge, grassland, road margin, and wetlands) to account for the effect of local cover independent of the surrounding landscape composition. 
Day of year was included as a 1-dimensional smooth, and easting and northing (km east and north of the trap extent centre point) were included as 2-dimensional smooths, in order to account for underlying spatial and temporal autocorrelation in the data.
The deviance residuals from each model were visually inspected for normality and equal variance [@hilbe2011; @wood2017].

@@

***
**Box 1: Interpretation of Functional Regression Plots**

This box explains how to interpret functional regression plots.

***

@@

Grassland, cereal, and canola were the three most-abundant landcover classes surrounding our traps, accounting for `r coverList %>% group_by(cover) %>% summarize(p=sum(prop)) %>% ungroup() %>% arrange(desc(p)) %>% mutate(prop=round(100*p/sum(p),2)) %>% slice(1:3) %>% pull(prop) %>% sum `% of land cover (Figure \@ref(fig:landscapeComp)).
Several landscape ``fingerprints" were evident in the landscape annuli, with cereal cover increasing with distance away, along with a corresponding decrease in canola cover.
Grassland cover was largely constant with distance, although a cluster of sites had uniformly high or low cover of grassland with distance.
The pitfall traps caught a total of `r with(Ncatches,total[spp=='PteMel'])` _Pterostichus melanarius_, `r with(Ncatches,total[spp=='ParDis'])` _Pardosa distincta_, `r with(Ncatches,total[spp=='ParMoe'])` _Pardosa moesta_, and `r with(Ncatches,total[spp=='Opilio'])` _Phalangium opilio_ (mean trapping rates per day: `r with(Ncatches,paste(round(perDayMean,2),collapse=', '))`, SD: `r with(Ncatches,paste(round(perDaySD,2),collapse=', '))`, respectively).

@@

<!-- Percent signs are escaped in LaTeX, but this causes havoc if used in fig.cap of chunk options (OK if used in internal captions, I think). Yihui uses "text references" instead. -->
(ref:PteMelCap) Landscape influence on _Pterostichus melanarius_ activity-density. Lines and dots represent means, and bars and shaded regions represent 95 % confidence intervals (1.96 x SE). Coloured regions represent early-, mid-, and late-season effects (red, purple, blue).

@@

(ref:ParDisCap) Landscape influence on _Pardosa distincta_ activity-density. Lines and dots represent means, and bars and shaded regions represent 95 % confidence intervals (1.96 x SE). Coloured regions represent early-, mid-, and late-season effects (red, purple, blue).

@@

(ref:ParMoeCap) Landscape influence on _Pardosa moesta_ activity-density. Lines and dots represent means, and bars and shaded regions represent 95 % confidence intervals (1.96 x SE). Coloured regions represent early-, mid-, and late-season effects (red, purple, blue).

@@

_Phalangium opilio_ activity-density was strongly influenced by trap location, with grassy field edges and wetlands having the highest activity-density (Figure \@ref(fig:OpilioLandscape)a).
Landscape composition explained `r filter(r2Nakagawa2,grepl('Opilio',model),!grepl('s\\(',terms),terms!='trapLoc',terms!='sigmaD') %>% summarize(s=sum(r2comp)) %>% pull(s) %>% paste(.,collapse=', ')`% of variance in _P. opilio_ activity-density.
Nearby grassland (p=`r with(opilSmTerms,p[Smoother=='ti(Distance,Day):GrassWetland'])`) and woodland (p=`r with(opilSmTerms,p[Smoother=='s(Distance):TreeShrub'])`) both had negative effects on _P. opilio_ activity-density, but only early in the season (Figures \@ref(fig:OpilioLandscape)b,c).
Compare this to _P. melanarius_, which is also influenced by grassland and woodland, but where the effect was negative _late_ in the season.
The temporal and spatial components, as in the other two species, were both very strong (both p`r with(opilSmTerms,p[grepl('E,N',Smoother)])`).
The landscape composition and trap location explained `r r2Nakagawa$OpilioMod[1]*100`%  of variance in activity-density, while the spatial and temporal smoothers accounted for `r with(r2Nakagawa,(OpilioMod[2]-OpilioMod[1])*100)`%.

(ref:OpilioCap) Landscape influence on _Phalangium opilio_ activity-density. Lines and dots represent means, and bars and shaded regions represent 95 % confidence intervals (1.96 x SE). Coloured regions represent early-, mid-, and late-season effects (red, purple, blue).

@@

r2Nakagawa2 %>% select(model,terms,r2comp) %>% mutate(terms=as.character(terms),r2comp=r2comp*1) %>%
  pivot_wider(names_from=model,values_from=r2comp) %>%
  mutate(terms=case_when(terms=='trapLoc' ~ 'Trap Location', terms=='s(day)' ~ 'Day of Year',
                         terms=='s(E,N)' ~ 'Spatial Location', terms=='GrassWetland' ~ 'Grassland',
                         terms=='TreeShrub' ~ 'Woodland', terms=='sigmaD' ~ 'Residual',
                         terms=='Urban' ~ 'Road margins', TRUE ~ terms)) %>%
  kable(.,format='latex',caption='R$^2_{GLMM}$ for model components, representing the proportion of total variance attributable to each term.',
        col.names=linebreak(x=c('Term','\\emph{Pterostichus}\n\\emph{melanarius}','\\emph{Pardosa}\n\\emph{distincta}',
                              '\\emph{Pardosa}\n\\emph{moesta}','\\emph{Phalangium}\n\\emph{opilio}'),align='l'),
        digits=c(1,3,3,3,3),align='rllll',escape=FALSE)

@@

_Phalangium opilio_ activity-density responded to grass and wetland cover, as well as trees and shrubs, but the trap location and landscape explained only `r r2Nakagawa$OpilioMod[1]*100`% of the variance.
Other studies have found that _P. opilio_ is mostly found in human-altered landscapes with a large proportion of farming [@muster2014; @vandepoel2015Msc].
Since most of our study area was in or near farmed land, the lack of variation is understandable, but there were a few notable landscape-level patterns that emerged.
_P. opilio_ activity-density was higher in grassy field edges and wetlands, but grasslands and wetlands acted as a local sink of _P. opilio_ until late in the season.
These indicate that while grass and wetlands could act as a reservoir, spillover into other land cover types may be limited, as _P. opilio_ avoids heavily grazed areas, possibly due to low humidity [@sajna2011].
_P. opilio_ is also found in large numbers at the edges of forests, and may migrate from forests into farmlands [@vandepoel2015Msc].
We found the opposite pattern: forests act as a weak sink for _P. opilio_, at least early in the year.
Unlike the other arthropods, most of the variation in activity-density for _P. opilio_ was explained by day of year (R$^2$ = `r round(filter(r2Nakagawa2,model=='OpilioMod',grepl('day',terms))$r2comp,2)`), indicating strong seasonal emergence cues.
<!-- Feels like this paragraph ends too soon. Why was there so little variation due to landscape? Link to biology of the critter. -->
